---
title: "Intro to time series wrangling and viz"
author: "Grace Brofman"
date: "1/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library(feasts) # for ACF
library(slider)

```

## Read in the data
```{r}

toolik <- read_csv(here("data", "toolikweather.csv")) #date is currently classified as a character (which is not good for time series)

```

### Convert this into a tsibble (time series tibble)

use lubridate to convert to a date class that r understands, then tsibble to convert to tsibble

```{r}
# using same column name on either side of mutate just overrides that column
# lubridate: enter date as it is first written so lubridate can understand (e.f. mdy, dmy, etc.)
toolik_ts <- toolik %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  as_tsibble(key = NULL, index = date)
# index is whatever column contains your time series information  
  
```

```{r}
ggplot(data = toolik_ts, 
       aes(x = date, y = mean_airtemp)) +
  geom_line()


```

## Use "index_by()" function to aggregate time series information by intervals we specify (equivalent of group_by() in dplyr)

```{r}

# calculate the mean monthly air temperature- for each year for each month
toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% #(.)split this up by year and month across entire time series
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))

ggplot(data = toolik_month,
       aes(x = yr_mo, y = monthly_mean_temp)) +
  geom_line() +
  facet_wrap(~month(yr_mo, label = TRUE)) #facet_wrap using lubridate

```
Aggregate by week:
```{r}
toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_airtemp = mean(mean_airtemp, na.rm = TRUE),
            min_airtemp = min(mean_airtemp))

```

## Filtering time series
Use filter_index() to specify ranges of time series to keep or exclude.
```{r}

# filter to only include observations from June 2000 - October 2001
toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10") # yyyy-mm

```
```{r}
toolik_ts %>% 
  filter_index("2006-04-10" ~ "2006-05-15")  # specifies range of dates
```

## Seasonplots

```{r}
toolik_ts %>% 
  filter(year(date) > 2014)  %>% # pull the year from the date column, keep obs > 2014
  gg_season(y = mean_airtemp)
# recognizes each year's season, then plots each year as a separate series

toolik_month %>% 
  gg_season(y = monthly_mean_temp) +
  theme_light()
```

## Seasonal subseries plot

```{r}
toolik_month %>% 
  gg_subseries(monthly_mean_temp)

```

## Calculating moving averages for time series data
Using the slider package in R
```{r}
# generating vector of random values
set.seed(2021)
test <- rnorm(n = 100, mean = 40, sd = 10)

slide(test, ~.x, .before = 2, .after = 2) # creates a window: for each observation, show that observation as well as the 2 before and the 2 after
# orignial output from slide is a list (as opposed to numeric)
```

```{r}
# moving average for windows of size 15 (days)
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp,
                                   mean,
                                   .before = 7,
                                   .after = 7)))

ggplot() +
  geom_line(data = toolik_ts, 
            aes(x = date, y = mean_airtemp), 
            size= 0.2, 
            color = "gray") +
  geom_line(data = roll_toolik_15,
            aes(x = date, 
                y = ma_15d)) +
  theme_minimal() 

```

## Create the ACF (autocorrelation function) visual represeentiation of how obs are correlated with prior observations on the same variable over specified lags (days, months, etc.)
```{r}
toolik_ts %>% 
  ACF(mean_airtemp) %>% 
  autoplot()
# daily: tells us that observations separated by 1 day are highly correlated, decreasing slightly with each day

toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot()
# we see that observations that are 12 months apart ten to be strongly positively correlated. but observations that are 6 or 18 months apart are strongly negatively correlated. dotted blue lines show significance. shows that there is not a lot of dampening happening.
```

## Time series decomposition by STL (not classical)

```{r}
toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf))) # start with monthly mean temperature modeled as a function of the different seasons

components(toolik_dec) %>% 
  autoplot()

# PANELS
#1. Original timeseries data
#2. Any trend that's pulled out
#3. Average seasonality
#4. Residual
# PAY ATTENTION TO SCALES!

```




